---
title: "temp"
author: "Abhraneel Sarma"
date: "2/19/2022"
output:
  html_document:
    css: ../inst/css/styles.css
  pdf_document: default
---

```{r setup, include=FALSE} 
library(dplyr)
library(tidyr)
library(purrr)
devtools::load_all(".")
```

<script src="https://code.jquery.com/jquery-3.6.0.min.js" integrity="sha256-/xUj+3OJU5yExlq6GSYGSHk7tPXikynS7ogEvDej/m4=" crossorigin="anonymous"></script>
<script src="./Tangle.js"></script>
<link rel="stylesheet" href="TangleKit/TangleKit.css" type="text/css">
<script type="text/javascript" src="TangleKit/mootools.js"></script>
<script type="text/javascript" src="TangleKit/sprintf.js"></script>
<script type="text/javascript" src="TangleKit/BVTouchable.js"></script>
<script type="text/javascript" src="TangleKit/TangleKit.js"></script>
<script>
  let choices, combos;
  let comboExists = (curr, combos) => {
    curr = JSON.stringify(curr);
    for (let combo of combos) {
      if (curr == JSON.stringify(combo)) return true;
    }
    return false;
  }

  Tangle.classes.Iterate = {
    initialize: function(element, options, tangle, variable) {
      element.addEvent("click", (event) => {
        choices = tangle.getValue("choices");
        combos = tangle.getValue("combos");
        let keys = Object.keys(choices);
        let curr = {}
        for (let key of keys) {
          curr[key] = tangle.getValue(key)
        }
        let it = 0;
        do {
          if (options.i === choices[variable].length-1)
            options.i = 0;
          else
            options.i++;
          curr[variable] = choices[variable][options.i];
        } while (!(comboExists(curr, combos)) && it++ < 4);
        tangle.setValue(variable, choices[variable][options.i]);
      })
    },
    update: function(element, value) {
      element.textContent = element.getAttribute("data-var") + ": " +value;
    }
  }

  function setup() {
    let elem = $("p#multiverse-params").get(0);
    let t = new Tangle(
      elem,
      {
        initialize: function() {
          const pre = $("pre.multiverse").get();
          this.choices = {};
          this.combos = [];
          for (let e of pre) {
            let paramOption = {};
            for (let className of e.classList) {
              if (className.includes("---")) {
                let split = className.split("---");
                paramOption[split[0]] = split[1];
                if (split[0] in this.choices) this.choices[split[0]].add(split[1]);
                else this.choices[split[0]] = new Set([split[1]]);
              }
            }
            this.combos.push(paramOption);
          }
          this.combos = new Set(this.combos);
          this.keys = Object.keys(this.choices);
          for (let k of this.keys) {
            this.choices[k] = [...this.choices[k]];
            this[k] = this.choices[k][0];
          }
        },
        update: function() {
          let universe = {}
          for (let k of this.keys) {
            universe[k] = this[k];
          }
          $("pre.multiverse").hide();
          $("." + Object.entries(universe).map(d => d.join('---')).join('.')).show();
        }
      }
    )
  }
  
  // given a specification of the multiverse
  // and a list of parameters
  // return a list of classess to be added
  function getClasses(spec, p) {
    let classes = Object.entries(spec).filter(i => !(p.includes(i[0])))
    return classes.map(d => d[1].map(x => d[0] + "---" + x)).flat()
  }
  
  const load = () => {
    const pre = $("pre.multiverse").get();
    let multiverse_spec = {}
    for (let e of pre) {
      let universe = [...e.classList].filter(i => i.includes("---")).map(d => d.split("---"));
      universe.forEach(d => {
        if (Object.keys(multiverse_spec).includes(d[0])) {
          if (!multiverse_spec[d[0]].includes(d[1])) {
            multiverse_spec[d[0]].push(d[1])
          }
        } else {
          multiverse_spec[d[0]] = [d[1]]
        }
      })
    }
    
    let parameters = Object.keys(multiverse_spec);
    // creates a tangle widget for each parameter in the declared multiverse
    for (let i of parameters) {
        $("p#multiverse-params").append(`<span class="Iterate TKSwitch" data-var=${i} data-i=0></span><br>`)
    }
    // adds classes for <pre> tags where parameters are missings
    for (let e of pre) {
      let current_params = [...e.classList].filter(i => i.includes("---")).map(d => d.split("---")[0]);
      // first get the elements where we need to add classes
      if (current_params.length != parameters.length) {
        // for each of the elements, add classes for each p---o combinations of the parameters not included
        let to_add = getClasses(multiverse_spec, current_params)
        e.classList.add(...to_add)
      }
    }
    
    setup();
    
    $("div.main-container").css("margin-top", 40 + $("div.multiverse-navbar").height());
  }
  
  window.onload = load;
</script>


<div class="multiverse-navbar">
  <p id="multiverse-params"></p>
</div>

## Multiverse case study #4

```{r}
data("hurricane")

hurricane_data <- hurricane %>%
    # rename some variables
    rename(
        year = Year,
        name = Name,
        dam = NDAM,
        death = alldeaths,
        female = Gender_MF,
        masfem = MasFem,
        category = Category,
        pressure = Minpressure_Updated_2014,
        wind = HighestWindSpeed
    ) %>%
    # create new variables
    mutate(
        post = ifelse(year>1979, 1, 0),
        zcat = as.numeric(scale(category)),
        zpressure = -scale(pressure),
        zwind = as.numeric(scale(wind)),
        z3 = as.numeric((zpressure + zcat + zwind) / 3)
    )
```


## Multiverse Analysis

To implement a multiverse analysis, we first need to create the `js parameter('masfem')` object:

```{r}
M = multiverse()
```

```{multiverse default-m-1, inside = M}
# here we just create the variable `df` in the multiverse
df = hurricane_data

# here, we perform a `filter` operation in the multiverse
df.filtered = df %>%
  filter(branch(death_outliers,
      "no_exclusion" ~ TRUE,
      "most_extreme" ~ name != "Katrina",
      "two_most_extreme" ~ !(name %in% c("Katrina", "Audrey"))
))
```

```{multiverse label = variable_definitions, inside = M}
df.filtered = df.filtered %>%
    mutate(
        femininity = branch(femininity_calculation,
          "masfem" ~ masfem,
          "female" ~ female
        ),
        damage = branch(damage_transform,
          "no_transform" ~ identity(dam),
          "log_transform" ~ log(dam)
        )
    )
```

```{multiverse default-m-2, inside = M}
mean(df.filtered$femininity)
```


```{multiverse default-m-5, inside = M}
hist(df.filtered$femininity)
```


```{r}
expand(M)
```


