---
title: "Integrating tangle.js with the output on multiverse notebook analysis"
output: html_document
---


```{r}
devtools::load_all(".")
```


```{css, echo=FALSE}
.main-container {
  margin-top: 48px;
}
p {
  font-size: 64px;
}
#asdf {
  margin-top: -16px;
  position: fixed;
  left: 0;
  top: 0;
}
span:active {
  user-select: none;
}
```

<script src="./Tangle.js"></script>
<link rel="stylesheet" href="TangleKit/TangleKit.css" type="text/css">
<script type="text/javascript" src="TangleKit/mootools.js"></script>
<script type="text/javascript" src="TangleKit/sprintf.js"></script>
<script type="text/javascript" src="TangleKit/BVTouchable.js"></script>
<script type="text/javascript" src="TangleKit/TangleKit.js"></script>
<script>
  let comboExists = (curr, combos) => {
    curr = JSON.stringify(curr);
    for (let combo of combos) {
      if (curr == JSON.stringify(combo)) return true;
    }
    return false;
  }

  Tangle.classes.Iterate = {
    initialize: function(element, options, tangle, variable) {
      element.addEvent("click", (event) => {
        let choices = tangle.getValue("choices");
        let combos = tangle.getValue("combos");
        console.log(choices)
        console.log(combos)
        let keys = Object.keys(choices);
        let curr = {}
        for (let key of keys) {
          curr[key] = tangle.getValue(key)
        }
        let it = 0;
        do {
          if (options.i === choices[variable].length-1)
            options.i = 0;
          else
            options.i++;
          curr[variable] = choices[variable][options.i];
        } while (!(comboExists(curr, combos)) && it++ < 4);
        tangle.setValue(variable, choices[variable][options.i]);
      })
    },
    update: function(element, value) {
      element.textContent = value;
    }
  }

  function setup() {
    <!-- depending on implementation, i think i might have to create the <span class="Iterate"> here, since it's kind of hard-coded right now. -->
    <!-- if i were to do this, i also have to do the this.data stuff here. -->
    var t = new Tangle(
      document.getElementById("asdf"),
      {
        initialize: function() {
          const pre = document.querySelectorAll("pre.multiverse");
          this.choices = {};
          this.combos = [];
          for (let e of pre) {
            let paramOption = {};
            for (let className of e.classList) {
              if (className.includes("---")) {
                let split = className.split("---");
                paramOption[split[0]] = split[1];
                if (split[0] in this.choices) this.choices[split[0]].add(split[1]);
                else this.choices[split[0]] = new Set([split[1]]);
              }
            }
            this.combos.push(paramOption);
          }
          this.combos = new Set(this.combos);
          this.keys = Object.keys(this.choices);
          for (let k of this.keys) {
            this.choices[k] = [...this.choices[k]];
            this[k] = this.choices[k][0];
          }
        },
        update: function() {
        }
      }
    )
  }
  document.body.onload = setup;
</script>


<p id="asdf">
  <span class="Iterate TKSwitch" data-var="a" data-i=0></span>
	<span class="Iterate TKSwitch" data-var="b" data-i=0></span>
	<span class="Iterate TKSwitch" data-var="c" data-i=0></span>
</p>
<div id="insert"></div>


```{r}
M = multiverse()
```


```{multiverse default-m-1, inside = M}
a = branch(a, 
           "1" ~ 1, 
           "2" ~ 2, 
           "3" ~ 3
          )
b = branch(b, 
           "x" %when% (a != 1) ~ "x", 
           "y" ~ "y", 
           "z" ~ "z"
          )
c = branch(c, 
           "4" ~ 4, 
           "5"%when% (b != "z") ~ 5
          )
```

<!-- 
∄(a==2 <=> b==y), ((c==4 && a==1) <=> b==x), (b==z <=> c==5)
<pre class='multiverse a---2 b---x c---4'>2x4</pre>
<pre class='multiverse a---3 b---x c---4'>3x4</pre>
<pre class='multiverse a---1 b---y c---4'>1y4</pre>
<pre class='multiverse a---3 b---y c---4'>3y4</pre>
<pre class='multiverse a---1 b---z c---4'>1z4</pre>
<pre class='multiverse a---2 b---z c---4'>2z4</pre>
<pre class='multiverse a---3 b---z c---4'>3z4</pre>
<pre class='multiverse a---1 b---x c---5'>1x5</pre>
<pre class='multiverse a---2 b---x c---5'>2x5</pre>
<pre class='multiverse a---3 b---x c---5'>3x5</pre>
<pre class='multiverse a---1 b---y c---5'>1y5</pre>
<pre class='multiverse a---3 b---y c---5'>3y5</pre>  -->
