---
title: "Exporting the output of a multiverse to JSON for Milliways, a multiverse visualisation tool"
author: "Abhraneel Sarma"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Exporting the output of a multiverse to JSON for Milliways, a multiverse visualisation tool}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(broom)
library(jsonlite)
library(distributional)
library(multiverse)
```

## Introduction

The correct interpretation of a multiverse analysis can be difficult due to their potential size and the complexity of correctly interpreting their uncertainty. Our recent work in developing [Milliways](https://abhsarma.github.io/milliways/), an interactive visualisation interface for the principled evaluation and interpretation of the results of multiverse analysis aims to address this problem. For more details, please refer to our [paper](https://mucollective.northwestern.edu/files/2024-milliways.pdf).

## Milliways

To visualise the results of a multiverse analysis using Milliways, the user needs to provide four files:

- `results`: a JSON file which contains the estimates from each universe in the multiverse analysis.
- `code`: a JSON file which contains the code used to generate the multiverse object.
- `data`: a JSON file which contains the dataset used in the multiverse analysis.
- `analysis`: an HTML file which contains the entire analysis as an Explorable Multiverse Analysis Report (EMAR).

In this document, I outline how to generate each of these files for the multiverse analysis on the Durante dataset (see `vignette("visualising-multiverse")`). Compiling this document will, by itself, result in the creation of the EMAR, provided the following function is declared.

Note #1: If compiling the document as an EMAR, each code chunk will execute automatically. You do not need to call `execute_multiverse()` to obtain the results unless you want to extract the results and access it from an R code chunk.

Note #2: If you are not compiling the document as an EMAR, the code chunks will not execute automatically. The only way to obtain the results is to first execute all the universes in the multiverse using `execute_multiverse()` and then accessing the results using an R code chunk.

Note #3: On the use of magrittr pipes (`%>%`) instead of the native R pipe (`|>`). Because `multiverse` rewrites R expressions, when I parse the code declared into individual R scripts, the native R pipe is "evaluated" (i.e. `df |> mutate(...) |> filter(...)` becomes `filter(mutate(df, ...), ...)`); this makes the code readable. Instead I use magrittr pipes which does not "evaluate" the R expressions.

```{r, eval = FALSE}
knit_as_emar()
```

## Analysis

The analysis follows the same steps outlined in `vignette("example-durante")`. The first step is to read the raw data from the file and store it as a tibble.

```{r cars}
data("durante")

data.raw.study2 <- durante %>%
  mutate(
    Abortion = abs(7 - Abortion) + 1,
    StemCell = abs(7 - StemCell) + 1,
    Marijuana = abs(7 - Marijuana) + 1,
    RichTax = abs(7 - RichTax) + 1,
    StLiving = abs(7 - StLiving) + 1,
    Profit = abs(7 - Profit) + 1,
    FiscConsComp = FreeMarket + PrivSocialSec + RichTax + StLiving + Profit,
    SocConsComp = Marriage + RestrictAbortion + Abortion + StemCell + Marijuana
  )

M = multiverse()
```


### Cycle Length

In their Durante et al. exclude participants based on the length of their menstrual cycle, and only include those whose cycle lengths are between 25 and 35 days. However, according to Steegen et al., due to the flexibility in the data collection, "this exclusion criterion can be instantiated in two reasonable ways, using either a woman’s computed cycle length or a woman’s self-reported typical cycle length."

Note: we can define a tangle widget to allow the user to switch between which operationalisation of the outlier exclusion criteria is used. Please look at the RMarkdown file to see how it is declared.

- <mv param="cycle_length"/>

```{multiverse default-m-1, inside = M}
df <- data.raw.study2 %>%
    mutate( ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast )  %>%
    filter( branch(cycle_length,
        "cl_option1" ~ TRUE,
        "cl_option2" ~ ComputedCycleLength > 25 & ComputedCycleLength < 35,
        "cl_option3" ~ ReportedCycleLength > 25 & ReportedCycleLength < 35
    ))
```

### Certainty 

Steegen et al. describe that how certain the participants are in their reported dates can be another justifiable exclusion criteria:

- <mv param="certainty"/>

```{multiverse default-m-2, inside = M}
df <- df %>%
    filter( branch(certainty,
        "cer_option1" ~ TRUE,
        "cer_option2" ~ Sure1 > 6 | Sure2 > 6
    ))
```

### Menstrual Calculation

The flexibility in how the data is collected also allows three reasonable alternatives for estimating a woman’s next menstrual onset, which is an intermediate step in determining cycle day.

- <mv param="menstrual_calculation"/>

```{multiverse default-m-3, inside = M}
df <- df %>%
    mutate(NextMenstrualOnset = branch(menstrual_calculation,
        "mc_option1" ~ StartDateofLastPeriod + ComputedCycleLength,
        "mc_option2" %when% (cycle_length != "cl_option2") ~ StartDateofLastPeriod + ReportedCycleLength,
        "mc_option3" ~ StartDateNext)
    )  %>%
    mutate(
      CycleDay = 28 - (NextMenstrualOnset - DateTesting),
      CycleDay = ifelse(WorkerID == 15, 11, ifelse(WorkerID == 16, 18, CycleDay)),
      CycleDay = ifelse(CycleDay > 1 & CycleDay < 28, CycleDay, ifelse(CycleDay < 1, 1, 28))
    )
``` 


### Fertility

Durante et al. classify women into a high or low fertility group based on cycle day, but this classification can be done in various different reasonable ways:

- <mv param="fertile"/>

```{multiverse default-m-4, inside = M}
df <- df %>%
    mutate( Fertility = branch( fertile,
        "fer_option1" ~ factor( ifelse(CycleDay >= 7 & CycleDay <= 14, "high", ifelse(CycleDay >= 17 & CycleDay <= 25, "low", NA)) ),
        "fer_option2" ~ factor( ifelse(CycleDay >= 6 & CycleDay <= 14, "high", ifelse(CycleDay >= 17 & CycleDay <= 27, "low", NA)) ),
        "fer_option3" ~ factor( ifelse(CycleDay >= 9 & CycleDay <= 17, "high", ifelse(CycleDay >= 18 & CycleDay <= 25, "low", NA)) ),
        "fer_option4" ~ factor( ifelse(CycleDay >= 8 & CycleDay <= 14, "high", "low") ),
        "fer_option5" ~ factor( ifelse(CycleDay >= 8 & CycleDay <= 17, "high", "low") )
    ))
```

### Relationship Status

The participants in the study described their relationship status as one of the following options: (1) not dating/romantically involved with anyone, (2) dating or involved
with only one partner, (3) engaged or living with my partner, and (4) married. This allows various different ways of classifying whether a participants is in a relationship or not:

- <mv param="relationship_status"/>

```{multiverse default-m-5, inside = M}
df <- df %>%
    mutate(RelationshipStatus = branch(relationship_status,
        "rs_option1" ~ factor(ifelse(Relationship==1 | Relationship==2, 'Single', 'Relationship')),
        "rs_option2" ~ factor(ifelse(Relationship==1, 'Single', 'Relationship')),
        "rs_option3" ~ factor(ifelse(Relationship==1, 'Single', ifelse(Relationship==3 | Relationship==4, 'Relationship', NA))) )
    ) %>%
  mutate( RelComp = round((Rel1 + Rel2 + Rel3)/3, 2))
```

## Regression Model

The authors perform an ANOVA to study the effect of *Fertility*, *Relationship* and their interaction term, on the composite Religiosity score. We fit the linear model using the call: `lm( RelComp ~ Fertility * RelationshipStatus, data = df )` inside our multiverse and save the result to a variable called `fit_RelComp`.

```{multiverse default-m-6, inside = M}
fit_RelComp <- lm(RelComp ~ Fertility * RelationshipStatus, data = df)
summary_RelComp <- fit_RelComp %>% 
    broom::tidy( conf.int = TRUE )

summary_RelComp
```


## Visualising the Results

```{multiverse default-m-7, inside = M}
p_val = round(summary(fit_RelComp)$coefficients[4, 4], 3)
if (p_val < 0.05) p_val = paste0(p_val, "*")

broom::augment(fit_RelComp, interval = "confidence") %>%
  group_by(Fertility, RelationshipStatus) %>%
  mutate(RelationshipStatus = ifelse(RelationshipStatus == "Relationship", "InRelationship", "Single")) %>%
  summarise(.fitted = mean(.fitted), .upper = mean(.upper), .lower = mean(.lower), .groups = "drop") %>%
  ggplot(aes(x = RelationshipStatus, y = .fitted, fill = Fertility)) + 
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single"), width = 0.5) + 
  geom_linerange(aes(ymin = .lower,  ymax = .upper), position = position_dodge(width = 0.5)) + 
  geom_text(label = paste0("Interaction:", p_val), x = 1.5, y = 7.5, colour = "#666666") + 
  labs(x = "RelationshipStatus", y = "ReligiosityCompositeScore") + 
  ylim(c(0, 8.1)) + 
  theme_minimal()
```



## Exporting the JSON files

The current section describes how the remaining three JSON files can be created. If you haven't already, the first step would be to execute all the analyses declared in the multiverse using `execute_multiverse()`:

```{r, eval = FALSE}
execute_multiverse(M)
```

We can then use dedicated functions for each of the JSON files: 
- `export_results_json` or `export_results_dist_json`.
- `export_code_json`

### Exporting the results

We provide two functions to export the results: `export_results_json` and `export_results_dist_json`. 
`export_results_json` requires the user to specify the following arguments:

- `term`: column name which contains the names of the outcome variables. For example, in the case of a regression, you could use the output of [broom::tidy()](https://generics.r-lib.org/reference/tidy.html) and thus the argument would be the column which contains the coefficient names.
- `estimate`: column name containing the mean / median point estimates for each outcome.
- `standard error`: column name containing the standard errors for each outcome.
- `dist`: column name containing distributional objects for each outcome variable. Optional **only** if `estimate` and `standard error` arguments are provided. See below for more details.
- (optional) `filename`: if specified, the function will create a file in the specified path; if not specified, it will return the dataframe as shown below:

```{r}
expand(M) %>%
  extract_variables(summary_RelComp) %>%
  unnest( cols = c(summary_RelComp) ) %>%
  mutate( term = recode( term, 
                 "RelationshipStatusSingle" = "Single",
                 "Fertilitylow:RelationshipStatusSingle" = "Single:Fertility_low"
  )) %>%
  export_results_json(term, estimate, std.error) |> 
  unnest(results) |> 
  select(.universe, term, estimate, std.error, cdf.x, cdf.y)
```

The resultant JSON file consists of a list of objects (where each object corresponds to one analysis in the multiverse). Within this object, the results attribute contains a(nother) list of objects corresponding to each outcome variable. For e.g., here we have four coefficients (see the results of the regression model), and thus the results attribute will contain four objects. Each object has the following attributes:

- `term`: name of the outcome variable
- `estimate`: mean / median point estimate i.e., $\mathbb{E}(\mu)$ for any parameter $\mu$.
- `std.error`: standard error for the point estimate i.e., $\sqrt{\text{var}(\mu)}$
- `cdf.x`: a list of quantiles
- `cdf.y`: a list of cumulative probability density estimates corresponding to the quantiles

In addition, it also contains the following attributes, but these are not currently used by Milliways:

- `statistic`
- `p.value`
- `conf.low`
- `conf.high`

### A Note on Distributional Assumptions

For simplicity, we assume that each of the outcome variables follow a normal distribution. However, this may not always be the case. In this case, we recommend that you should specify the `dist` argument to `export_results_json` or use the `export_results_dist_json` which allows you to specify [distributional](https://pkg.mitchelloharawild.com/distributional/) objects for each outcome. We demonstrate how a user can do this with the following example of a multiverse analysis where the results consists of two parameters: $\mu \sim \text{N}(0, 1)$, a normally distributed random variable and $\sigma \sim \text{exp}(1)$, a random variable which follows the exponential distribution.

```{r}
expand_grid(
  .universe = seq(1:5),
  nesting(
    term = c("mu", "sigma"),
    dist = c(dist_normal(0, 1), dist_exponential(1))
  )
) |> 
  export_results_dist_json(term, dist) |> 
  unnest(results)
```

### Exporting the Code

Exporting the code file is relatively simple as the only arguments that need to be provided are the multiverse object and file path:

```{r, eval = FALSE}
export_code_json(M, "code.json")
```

The JSON file consists of two attributes: `code` and `parameters`. 
`code` is a list of strings consisting of the R and multiverse syntax used to implement the analysis. For readability, we use [styler](https://styler.r-lib.org/) to break up the declared code. `parameters` is an object listing the parameter names and the corresponding options for each of the parameters declared in the analysis.


### Exporting the Dataset 

This function is used to export the (unmodified) dataset that is used in the analysis, and is a simple wrapper around the `write_json` function.

```{r, eval = FALSE}
 export_data_json(durante, "../../data.json")
```

The JSON file consists of a list of objects, each with two attributes: `field` and `values`. 
`field` is the name of a column corresponding to a variable in the dataset. `values` are a list of values for that variable in the dataset.

