---
title: "Creating a multiverse analysis"
output: html_document
---

## Introduction

In this document, we outline an initial approach to conducting a multiverse analysis in R. 

Data analysis can involve several decisions involving two or more options. In most statistical analysis, these decisions are taken by the researcher based on some reasonable justification. However, for several decisions, there can be more than one reasonable options to choose from. A multiverse analysis is a form of analysis which makes all such decisions explicit and conducts the complete analysis for all combinations of options (of each decision). 

Below, we illustrate an example of a single analysis for a dataset. And then extend it to a multiverse analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(tidyverse)
library(magrittr)
#library(multidy)
```

## The data
The first step is to read the raw data from the file and store it as a tibble. We will be following the *_tidy data_* format here. The data is stored in two text files, and we can use `readr` to read the files into R. In this example, we will use the data collected by *Durante et al.*, which investigated the effect of fertility on religiosity and political attitudes. We will focus on their second study (which we store in `data.raw.study2`).

```{r data}
data.raw.study2 <- read_delim("../data/durante_etal_2013_study2.txt", delim = "\t",
    col_types = cols(
        DateTesting = col_date(format = "%m/%d/%y"),
        StartDateofLastPeriod =col_date(format = "%m/%d/%y"),
        StartDateofPeriodBeforeLast = col_date(format = "%m/%d/%y"),
        StartDateNext = col_date(format = "%m/%d/%y")
  )) %>%
  mutate(
    Abortion = abs(7 - Abortion) + 1,
    StemCell = abs(7 - StemCell) + 1,
    Marijuana = abs(7 - Marijuana) + 1,
    RichTax = abs(7 - RichTax) + 1,
    StLiving = abs(7 - StLiving) + 1,
    Profit = abs(7 - Profit) + 1,
    FiscConsComp = FreeMarket + PrivSocialSec + RichTax + StLiving + Profit,
    SocConsComp = Marriage + RestrictAbortion + Abortion + StemCell + Marijuana
  ) %>%
  mutate(ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast ) 
```

The data look like this:

```{r}
data.raw.study2 %>%
  head(10)
```

The original paper looked at the relationship between fertility, relationship status, and religiousity. But there are many reasonable ways to have defined each of these three variables from this dataset, so it is a good candidate for multiverse analysis.

## A single data set analysis: one possible analysis among many

The data collected needs to be processed before it can be modeled. Preparing the data set for analysis can involve several steps and decisions regarding how to encode the different raw values. The following is one example of data processing that can be performed for this study. 

```{r single_analysis}
one_universe = data.raw.study2 %>%
  mutate( ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast ) %>%
  mutate( NextMenstrualOnset = StartDateofLastPeriod + ComputedCycleLength ) %>%
  mutate(
    CycleDay = 28 - (NextMenstrualOnset - DateTesting),
    CycleDay = ifelse(CycleDay > 1 & CycleDay < 28, CycleDay, ifelse(CycleDay < 1, 1, 28))
  ) %>%
  mutate(
    Relationship = factor(ifelse(Relationship==1 | Relationship==2, "Single", "Relationship"))
  ) %>%
  filter( ComputedCycleLength > 25 & ComputedCycleLength < 35) %>%
  filter( Sure1 > 6 | Sure2 > 6 ) %>%
  mutate( Fertility = factor( ifelse(CycleDay >= 7 & CycleDay <= 14, "high", ifelse(CycleDay >= 17 & CycleDay <= 25, "low", "medium")) ) )
```

The transformed data for this one universe looks like this:

```{r}
one_universe %>%
  select( NextMenstrualOnset, Relationship, Sure1, Sure2, Fertility, everything() ) %>%
  head(10)
```

```{r}
one_universe %>%
  ggplot(aes(x = Relationship, y = Rel1 + Rel2 + Rel3, color = Fertility)) +
  stat_summary(position = position_dodge(width = .1), fun.data = "mean_se")
```

However, there also exists other valid processing options: instead of calculating `NextMenstrualOnset = StartDateofLastPeriod + ComputedCycleLength`, it can also be calculated as `StartDateofLastPeriod + ReportedCycleLength`. Such alternate processing options can exist for several decisions that a researcher makes in the data processing, analysis and presentation stages. This can thus result in a *multiverse of analysis*, with the one described above representing a single *universe*. 

Below, we describe how our package allows you to conduct a multiverse analysis with ease.

## Defining the multiverse

`multidy` provides flexible functions which can be used to perform a multiverse analysis.

The first step is to define a *new multiverse*. We will use the multiverse object to create a set of universes, each representing a different way of analysing our data.

```{r}
M <- new("multiverse")
```

The next step is to define our possible analyses inside the multiverse. The `multidy` package includes functions that aim to make it easy to write multiverse analyses in as close a way to a single universe analysis as possible (as seen in the single analysis shown above). 

Consider these first few lines from the transformation code in the single analysis above:

```{r eval=FALSE}
df <- data.raw.study2 %>%
  mutate(ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast) %>%
  mutate(NextMenstrualOnset = StartDateofLastPeriod + ComputedCycleLength)
```

But `NextMenstrualOnset` could be calculated in at least two other reasonable ways:

* `NextMenstrualOnset = StartDateofLastPeriod + ReportedCycleLength`
* `NextMenstrualOnset = StartDateNext`

To create a multiverse that includes these three possible processing options, we can use the `branch()` function. The `branch()` function defines a *parameter* (here `menstrual_calculation`) and the different *options* that the parameter can take (here, `"mc_option1"`, `"mc_option2"`, `"mc_option3"`). Each option corresponds to a different chunk of code that would be executed in a different universe.

```{r parameter_options, eval = FALSE}
NextMenstrualOnset = branch(menstrual_calculation, 
  "mc_option1" ~ StartDateofLastPeriod + ComputedCycleLength,
  "mc_option2" ~ StartDateofLastPeriod + ReportedCycleLength,
  "mc_option3" ~ StartDateNext
)
```

The `branch()` function indicates that, *in our multiverse*, NextMenstrualOnset can take either of the three options (here, `"mc_option1"`, `"mc_option2"`, `"mc_option3"`). Thus, we need to declare this data processing step inside the multiverse. We do this by using the `inside()` function. The `inside()` function takes in two arguments:
1. the multiverse object, M; and
2. the code for the analysis (including branches), enclosed within `{}`

```{r add_to_multiverse}
M = inside(M, {
  df <- data.raw.study2  %>%
    mutate( ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast ) %>%
    mutate( NextMenstrualOnset = branch(menstrual_calculation, 
      "mc_option1" ~ StartDateofLastPeriod + ComputedCycleLength,
      "mc_option2" ~ StartDateofLastPeriod + ReportedCycleLength,
      "mc_option3" ~ StartDateNext)
    )
})
```

Alternative to the `inside` function, we can use the more concise `<-` operator to define the multiverse. However, any expression on the right-hand-side of the `<-` operator needs to be preceded by the `~` (tilde) which converts the expression to a formula. Thus, the syntax looks like: `<multiverse_object>$<variable_name> <- ~ <expression>`

```{r alternate_add_to_multiverse, eval = FALSE}
M$df <- ~ data.raw.study2  %>%
  mutate( ComputedCycleLength = StartDateofLastPeriod - StartDateofPeriodBeforeLast ) %>%
    mutate(NextMenstrualOnset = branch(mentrual_calculation, 
      "mc_option1" ~ StartDateofLastPeriod + ComputedCycleLength,
      "mc_option2" ~ StartDateofLastPeriod + ReportedCycleLength,
      "mc_option3" ~ StartDateNext)
    )
```

## Parsing the multiverse
The next step is to parse the multiverse to identify the `parameters` and the corresponding `options` that have been defined for each parameter. For this, we use the `parse_multiverse()` function.

```{r parse_multiverse, include=FALSE}
M.parsed = parse_multiverse(M)
```

The `multiverse` object we have defined consists of the following attributes: 
1. `parameters`, which is a list of parameters
```{r parameter_list}
M.parsed@parameters
```

2. `conditions`, which is a list of conditions (we'll define this later) _HAS NOT BEEN IMPLEMENTED YET_
3. `multiverse_table`, which is a tibble consisting of all possible combination of values for the multiverse
```{r}
M.parsed@multiverse_table
```

4. `current_parameter_assignment`, which is a list of one set of options for each parameter defined in the multiverse. It initialises to the first option for each parameter. This should be the default analysis.
```{r}
M.parsed$current_parameter_assignment
```

5. `code`, which is the code that the user passes to the multiverse to conduct a multiverse analysis. However, we do not execute this code and it is stored unevaluated. The user can interactively edit and rewrte this code, and can execute it for the current analysis or the entire multiverse using dedicated functions.
```{r}
M.parsed$code
```

## Running a single analysis from the multiverse

At this point, we have defined three possible processing options (three universes) in our multiverse, but no code has been executed. We need to be able to analyse each universe in turn. We can generate the code for a particular universe using the `get_code()`, and execute this code using `rlang::eval_tidy()`.

```{r generate_code}
execute_default(M.parsed) %>%
  select( NextMenstrualOnset, everything() ) %>%
  head()
```

These attributes are populated only after the `parse_multiverse()` function has been called on the multiverse object.

## A complete multiverse analysis
Besides calculating the onset of the next menstruation cycle, there are other variables which have multiple valid and reasonable processing options. These include defining `Relationship` and `Fertility`, and exclusion criteria based on the values for cycle length and certainty of responses. The next code chunk illustrates how this can be added to the multiverse object defined above.

```{r}
M = inside(M, {
  df %>%
    mutate(Relationship = branch( relationship_status, 
      "rs_option1" ~ factor(ifelse(Relationship==1 | Relationship==2, 'Single', 'Relationship')),
      "rs_option2" ~ factor(ifelse(Relationship==1, 'Single', 'Relationship')),
      "rs_option3" ~ factor(ifelse(Relationship==1, 'Single', ifelse(Relationship==3 | Relationship==4, 'Relationship', NA))) )
    ) %>%
    mutate(
      CycleDay = 28 - (NextMenstrualOnset - DateTesting),
      CycleDay = ifelse(CycleDay > 1 & CycleDay < 28, CycleDay, ifelse(CycleDay < 1, 1, 28))
    ) %>%
    filter( branch(cycle_length, 
      "cl_option1" ~ TRUE,
      "cl_option2" ~ ComputedCycleLength > 25 & ComputedCycleLength < 35,
      "cl_option3" ~ ReportedCycleLength > 25 & ReportedCycleLength < 35
    )) %>%
    filter( branch(certainty,
        "cer_option1" ~ TRUE,
        "cer_option2" ~ Sure1 > 6 | Sure2 > 6
    )) %>%
    mutate( Fertility = branch( fertile,
        "fer_option1" ~ factor( ifelse(CycleDay >= 7 & CycleDay <= 14, "high", ifelse(CycleDay >= 17 & CycleDay <= 25, "low", "medium")) ),
        "fer_option2" ~ factor( ifelse(CycleDay >= 6 & CycleDay <= 14, "high", ifelse(CycleDay >= 17 & CycleDay <= 27, "low", "medium")) ),
        "fer_option3" ~ factor( ifelse(CycleDay >= 9 & CycleDay <= 17, "high", ifelse(CycleDay >= 18 & CycleDay <= 25, "low", "medium")) ),
        "fer_option4" ~ factor( ifelse(CycleDay >= 8 & CycleDay <= 17, "high", "low") )
    ))
})
```

Since the multiverse object has already been created and the one parameter has already been defined, the inside function will add to the previous code.

```{r}
M@code
```

Next, we use the `parse_multiverse()` function

```{r, include = FALSE}
M.parsed = parse_multiverse(M)
```

We can then inspect the `multiverse_table` to see if all the possible combinations of the parameter options have been identified.

```{r}
M.parsed@multiverse_table %>%
  head(10)
```

Finally, we can execute the default single universe analysis from this multiverse:

```{r, eval = FALSE}
M.parsed %>%
  execute_default() %>%
  select(ComputedCycleLength, NextMenstrualOnset, CycleDay, Relationship, Fertility, everything()) %>%
  head()
```

Or, we could get the code for a different single universe analysis by providing a list of parameter assignments (options which each parameter should take) to the `get_code()` function and execute it:

```{r, eval = FALSE}
universe_parameter_assignment <- list(
    menstrual_calculation = "mc_option1",
    relationship_status = "rs_option3",
    cycle_length = "cl_option2",
    certainty = "cer_option1",
    fertile = "fer_option4"
  )

M.parsed %>%
  execute_default(universe_parameter_assignment)  %>%
  select(ComputedCycleLength, NextMenstrualOnset, CycleDay, Relationship, Fertility, everything()) %>%
  head()
```

