---
title: "A multiverse analysis of the efficiency of physical visualizations"
author: "Abhraneel Sarma"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A multiverse analysis of the efficiency of physical visualizations}
  %\VignetteEncoding{UTF-8}
---

## Introduction
```{r setup, include=FALSE}
library(knitr)
opts_knit$set(
  echo = TRUE,
  fig.width = 6, 
  fig.height = 4,
  output.dir="vignettes"
)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(broom)
library(gganimate)
library(multidy)
library(boot)
```


## The data

```{r}
data("userlogs")
data.userlogs.raw = userlogs
```


## Average task completion time (arithmetic mean) for each condition. 
Error bars are 95% BCa bootstrap CIs.

```{r}
M = multiverse()
inside(M, x <- 1)
inside(M, y <- 1)
code(M)
```


```{r}
samplemean <- function(x, d) {return(mean(x[d]))}

M = multiverse()

M$df <- ~ data.userlogs.raw %>%
  mutate( duration = branch( data_transform, 
                         "none" ~ duration,
                         "log" ~ log(duration)
        )) %>% 
  mutate( modality.f = factor(modality) ) %>%
  group_by(subject, modality.f, modalityname) %>% 
  summarise( duration = mean(duration) ) %>%
  group_by(modality.f) %>%
  nest( duration )

M$n_comparisons <- ~ 3

M$df <- ~ df %>%
  mutate( ci = branch( confidence_interval, 0.5, 0.68, 0.8, 0.9, 0.95, 0.99, 0.999 )) %>%
  mutate( fit = branch(bootstrap,
                       "t-dist" ~ map2(data, ci, ~ tidy(t.test(.x$duration, conf.level = .y)) ),
                       "bca" ~ map2(data, ci, ~ boot(data = .x$duration, statistic = samplemean, R = 5000) %>%
                              tidy( conf.int = TRUE, conf.method = "bca", conf.level = .y) %>%
                              rename(estimate = statistic))
        )) %>%
  select(-data) %>%
  unnest( fit )
```

```{r}
multiverse_table(M)
```


```{r, warnings = FALSE}
execute_multiverse(M)
```


```{r}
p <- multiverse_table(M) %>%
  unnest( map(.results, "df" ) )  %>%
  mutate( modality.f = recode( modality.f, 
                 "1" = "physical-touch",
                 "2" = "no-touch",
                 "3" = "virtual-prop",
                 "4" = "virtual-mouse"
  ) ) %>%
  mutate( 
    estimate = ifelse(data_transform == "log", exp(estimate), estimate),
    conf.low = ifelse(data_transform == "log", exp(conf.low), conf.low),
    conf.high = ifelse(data_transform == "log", exp(conf.high), conf.high)
  ) %>%
  ggplot() + 
  geom_vline( xintercept = 0,  colour = '#979797' ) +
  geom_point( aes(x = estimate, y = modality.f)) +
  geom_errorbarh( aes(xmin = conf.low, xmax = conf.high, y = modality.f), height = 0)  +
  transition_manual( index )

animate(p, nframes = 28,  fps = 2)
```

## Differences between mean completion times 
Differences between mean completion times (arithmetic means) between conditions. Error bars are 95% BCa bootstrap CIs.

```{r}
M$df.diffs <- ~ data.userlogs.raw  %>%
  mutate( duration = branch( data_transform, 
                         "none" ~ duration,
                         "log" ~ log(duration)
  )) %>%
  arrange(modality) %>%
  mutate(index = rep(seq(1:(nrow(.)/4)), 4)) %>%
  select(index, modalityname, duration) %>%
  spread(modalityname, duration) %>%
  mutate(
    `physical-touch - physical-notouch` = `physical-touch` - `physical-notouch`,
    `physical-notouch - virtual-prop` = `physical-notouch` - `virtual-prop`,
    `virtual-prop - virtual-mouse` = `virtual-prop` - `virtual-mouse`
  ) %>%
  select(`physical-touch - physical-notouch`, `physical-notouch - virtual-prop`, `virtual-prop - virtual-mouse`) %>%
  gather(levels) %>%
  group_by(levels) %>%
  nest()

M$df.diffs <- ~ df.diffs  %>%
  mutate( ci = branch( confidence_interval, 0.5, 0.68, 0.8, 0.9, 0.95, 0.99, 0.999 )) %>%
  mutate( diff = branch(bootstrap,
                       "t-dist" ~ map2(data, ci, ~ tidy(t.test(.x$value, conf.level = .y)) ),
                       "bca" ~ map2(data, ci, ~ boot(data = .x$value, statistic = samplemean, R = 5000) %>%
                              tidy( conf.int = TRUE, conf.method = "bca", conf.level = .y) %>%
                              rename(estimate = statistic))
        )) %>%
  unnest(diff)
```

```{r, warnings = FALSE}
execute_multiverse(M)
```

```{r}
M$df.diffs
```

```{r}
p <- multiverse_table(M) %>%
  unnest( map(.results, "df.diffs" ) ) %>%
  ggplot() + 
  geom_vline( xintercept = 0,  colour = '#979797' ) +
  geom_point( aes(x = estimate, y = levels)) +
  geom_errorbarh( aes(xmin = conf.low, xmax = conf.high, y = levels), height = 0) +
  transition_manual( index )

animate(p, nframes = 28,  fps = 2)
```


## Not relevant....Some Checks
Not sure why the std error is constant across levels. Figure out a way to have non-constant variance across levels

```{r, echo = FALSE, eval = FALSE}
fit <- df.s  %>% 
  mutate(
    time.exp = exp(logtime)
  ) %>%
  select( logtime, time.exp, modality.f ) %>%
  # filter( modality.f == 1) %>%
  lm(logtime ~ modality.f - 1, data = .)

broom::tidy(fit, conf.int = TRUE) %>%
  mutate( 
    time.conf.low = exp(conf.low),
    time.conf.high = exp(conf.high),
    time.mean = exp(estimate) 
  )
```


```{r, echo = FALSE, eval = FALSE}
df.s  %>%
  ungroup() %>%
  select( subject, time, modality.f ) %>%
  # filter( modality.f == 1) %>%
  lm(time ~ modality.f - 1, data = .) %>%
  broom::tidy(conf.int = TRUE)
```


